An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-10 Notes/BoundEffRed.tex
code and tex

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{euscript}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
%\usepackage{subfig}
\usepackage{color}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{dsfont}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathpazo}
\usepackage{array}

@@

\title{An Efficient Forecasting Approach for the Reduction of Boundary Effects in Time-Frequency Representations}

@@

Various improvements can be considered to make the algorithm more robust to certain situations. In particular, we have noticed that when the observed signal does not oscillate {\color{red}[HT: we need to confirm this statement]}, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to select the portions without sufficient quality, in order to reject the portions for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

>>>>

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{euscript}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
%\usepackage{subfig}
\usepackage{color}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{dsfont}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathpazo}
\usepackage{array}
\usepackage{bigints}

@@

\title{An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations}

@@

Various improvements can be considered to make the algorithm more robust to certain situations. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to select the portions without sufficient quality, in order to reject the portions for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

### 2020-07-10 Notes/algorithm.tex
code and tex

%
The pseudo-code of the proposed algorithm to reduce boundary effects on windowing-based TF representations is shown in the Algorithm~\ref{alg:boundary}. We coined the algorithm {\sf BoundEffRed}. Below, we detail the algorithm, particularly the Signal extension algorithm {\sf SigExt} in Algorithm~\ref{alg:boundary}.

\begin{algorithm}
\caption{Tackling boundary effects. $\bF_\bx = \mbox{\sf BoundEffRed}(\bx,M,K,L,\ccF)$}
\label{alg:boundary}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $L$, $\ccF$
\STATE 
\STATE {\bf Forecasting step.}
\STATE \quad\textbullet\ Signal extension: $\tilde\bx = \mbox{\sf SigExt}(\bx)$.
\STATE 
\STATE {\bf Representation step.}
\STATE \quad\textbullet\ Representation evaluation: $\ccF_{N+2L}(\tilde\bx)$.
\STATE \quad\textbullet\ Restriction of $\ccF_{N+2L}(\tilde\bx)$ to the central time interval (see~\eqref{eq:restriction}) to obtain $\bF_\bx=\ccF_N^{\mathrm{ext}}(\bx)$.
\STATE 
\STATE {\bf Output}: Signal representation $\bF_\bx$
\end{algorithmic}
\end{algorithm}

@@

Let $\ccF_{N}:\RR^{N}\to\RR^{F\times N}$ generically denotes the TF or time-scale representation we are interested in, which could be, for instance, STFT, CWT, SST, or RS. Here, $F$ typically denotes the size of the discretization along the frequency axis. Due to the boundary effects, the representation $\ccF_{N}(\bx)$ shows undesired patterns when approaching its edges. %For example, the instantaneous frequencies highlighted by the SST can be blurred near that edges. 
To alleviate the boundary effect, we apply the representation to the estimated extended signal $\tilde\bx$. This strategy moves the boundary effects out of the time interval $I=[0, \frac{N-1}{\fs}]$. Finally, the boundary-effects insensitive representation $\ccF_N^{\mathrm{ext}}:\RR^{N}\to\RR^{F\times N}$ of $\bx$ is given for all $\nu\in\{0,\cdots,F-1\}$, $n\in\{0,\cdots,N-1\}$ by:
\begin{equation}
\ccF_N^{\mathrm{ext}}(\bx)[\nu,n] = \ccF_{N+2L}(\tilde \bx)[\nu,L+n]\ .
\label{eq:restriction}
\end{equation}
This amounts to restricting the representation $\ccF_{N+2L}(\tilde \bx)$ to the original measurement interval of $\bx$. For the sake of simplicity, we denotes the restriction operator by $\cR$, where $\cR:\RR^{F\times (N+2L)}\to\RR^{F\times N}$. Consequently, we have:
\begin{equation*}
\ccF_N^{\mathrm{ext}}(\bx) = \cR\left( \ccF_{N+2L}(\tilde \bx) \right) \ .
\end{equation*}
We call $\ccF_N^{\mathrm{ext}}$ the {\em boundary-free TF representation}.

>>>>

%
The pseudo-code of the proposed algorithm to reduce boundary effects on windowing-based TF representations is shown in the Algorithm~\ref{alg:boundary}. We coined the algorithm {\sf BoundEffRed}. Below, we detail the algorithm, particularly the Signal extension algorithm {\sf SigExt} in Algorithm~\ref{alg:extension}.

\begin{algorithm}
\caption{Tackling boundary effects of a TF representation in real-time. $\bF_\bx = \mbox{\sf BoundEffRed}(\bx,M,K,L,\ccF)$}
\label{alg:boundary}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $N_0$, $L$, $\ccF$
\STATE
\WHILE {$N$ increases}
\STATE {\bf Forecasting step.}
\STATE \quad\textbullet\ Signal extension: $\tilde\bx = \mbox{\sf SigExt}(\bx_N)$.
\STATE 
\STATE {\bf Representation step.}
\STATE \quad\textbullet\ Representation evaluation: $\ccF_{N+2L}(\tilde\bx)$.
\STATE \quad\textbullet\ Restriction of $\ccF_{N+2L}(\tilde\bx)$ to the central time interval (see~\eqref{eq:restriction}) to obtain $\bF_\bx^{(N)}=\ccF_N^{\mathrm{ext}}(\bx_N)$.
\STATE 
\STATE {\bf Real-time output}: Signal representation $\bF_\bx^{(N)}$
\STATE $N \leftarrow N+1$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

@@

Let $\ccF_{N}:\RR^{N}\to\RR^{F\times N}$ generically denotes the TF or time-scale representation we are interested in, which could be, for instance, STFT, CWT, SST, or RS. Here, $F$ typically denotes the size of the discretization along the frequency axis. Due to the boundary effects, the representation $\ccF_{N}(\bx)$ shows undesired patterns when approaching its edges. %For example, the instantaneous frequencies highlighted by the SST can be blurred near that edges. 
To alleviate the boundary effect, we apply the representation to the estimated extended signal $\tilde\bx$. This strategy moves the boundary effects out of the time interval $I=[0, \frac{N-1}{\fs}]$. Finally, the boundary-effects insensitive representation $\ccF_N^{\mathrm{ext}}:\RR^{N}\to\RR^{F\times N}$ of $\bx$ is given for all $\nu\in\{0,\cdots,F-1\}$, $n\in\{0,\cdots,N-1\}$ by:
\begin{equation}
\ccF_N^{\mathrm{ext}}(\bx)[\nu,n] = \ccF_{N+2L}(\tilde \bx)[\nu,L+n]\ .
\label{eq:restriction}
\end{equation}
This amounts to restricting the representation $\ccF_{N+2L}(\tilde \bx)$ to the original measurement interval of $\bx$. For the sake of simplicity, we denotes the restriction operator by $\cR$, where $\cR:\RR^{F\times (N+2L)}\to\RR^{F\times N}$. Consequently, we have:
\begin{equation}
\ccF_N^{\mathrm{ext}}(\bx) = \cR\left( \ccF_{N+2L}(\tilde \bx) \right) \ .
\label{eq:bound.free.TFR}
\end{equation}
We call $\ccF_N^{\mathrm{ext}}$ the {\em boundary-free TF representation}.

### 2020-07-10 Notes/results.tex
code and tex

On Fig.~\ref{fig:res.noise.sine}, we display the experimental variance $\gamma_{\xp}[N-1+\ell,N-1+\ell]$ for each value of $\sigma$. The associated theoretical asymptotic forecasting variance~\eqref{eq:cov.error.2} is also displayed in solid line. As expected, this result highlights the fact that the forecasting variance increases linearly with respect to $\sigma^2$. Surprisingly, this result shows that the forecasting variance {\em slightly} decreases with $\ell$, what is counterintuitive. 
%
{\color{red} [HT: need a more convincing explanation.] It should be noted that, contrary to what expression~\eqref{eq:cov.error.2} suggests, smaller values of $\sigma$ do not cause a decrease of the experimental variance, but an increase. [HT: strange.]} This comes from the fact that when $\sigma$ is small, the matrix $\bX\bX^T$ becomes ill-conditioned. The calculation of the forecasting matrix $\tilde\bA$ in~\eqref{eq:lse} is then strongly disturbed.
\begin{figure}
%\includegraphics[width=.24\textwidth]{biasNoiseSine.eps}
\includegraphics[width=.48\textwidth]{VarianceNoiseSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $\sigma$.}
\label{fig:res.noise.sine}
\end{figure}

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary over time. The deterministic component $\bz$ of the random vector $\bx$ (constructed following the model~\eqref{eq:model.noise}) takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the additive noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$. Numerically, we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

To quantify the global quality (\ie~not depending on $\ell$) of the forecasting approaches, we evaluate the Experimental Mean Square Error $\mathrm{MSE_{xp}}(\tilde\bx)$ of the forward forecast extended signals, namely:
\begin{align}
\label{eq:mse}
\mathrm{MSE_{\xp}}(\tilde\bx) &= \dfrac1{L}\|\tilde\bx -\bx^\mathrm{ext}\|^2 \\
\nonumber
&=\! \dfrac1{L}\sum_{\ell=1}^L \bmu_{\xp}[N\!-\!1\!+\!\ell]^2 \!+\! \bgamma_{\xp}[N\!-\!1\!+\!\ell,N\!-\!1\!+\!\ell] .
\end{align}
where $\bx^\mathrm{ext}$ is the ground-truth extended signal, that is: $\bx^\mathrm{ext} = \begin{pmatrix}\bx[-L] & \cdots & \bx[N-1+L] \end{pmatrix}$. Then, as long as the bias $\bmu[N-1+\ell]$ and the variance $\bgamma[N-1+\ell,N-1+\ell]$ of the forecasting estimator remain small for all $\ell$, the MSE takes small values either. Corresponding results are given in Table~\ref{tab:mse.sine}. They show that the naive extension we propose gives satisfying results, in particular in comparison with the point-symmetric extension. Besides, even though the other more sophisticated methods, like GPR, give MSE values that have a slightly smaller standard deviation, these methods are substantially limited by the computing time they require, which prevent them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimize the trade-off between the forecasting quality and the computing time. %That is why, it is implemented in our algorithm for the reduction of boundary effects.

@@

\subsubsection{Metrics}
The quality of the boundary effects reduction must be evaluated directly on the TF representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$. The optimal representation is defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation*}
\ccF_N^\mathrm{opt}(\bx) = \cR\left( \ccF_{N+2L}(\bx^\mathrm{ext}) \right) \ .
\end{equation*}

In the aim of comparing the different techniques, we use a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given TF representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following probability density function:
\begin{align}
p_\ccQ^t(\xi) &= \dfrac{|\ccQ(\xi,t)|^2}{\int_\RR |\ccQ(\nu,t)|^2\dd\nu}\ .
\label{eq:ppdf}
\end{align}
Then, at each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$.
Finally, the distance between the two TF representations is obtained by averaging all the optimal transport distances with respect to time:
\begin{equation}
D(\ccQ,\ccF_0) = \frac1{|I|}\int_I d_{t}(\ccQ,\ccF_0)\dd t\ .
\end{equation}
The \textit{Optimal Transport Distance} (OTD) quantifies the proximity between the estimated and actual instantaneous frequencies while favouring the sparsity of the estimated TF representation.

Let us evaluate the quality of the boundary effects reduction on biomedical signals.

\subsubsection{Respiratory signal}
We first consider a 50 minutes-long respiratory signal. This signal is sampled at $\fs=100$~Hz. A zoom on a small portion of the signal is displayed in Fig.~\ref{fig:tho}.

@@

From that large signal, we build a dataset of $48$ non-overlapping signals of 60 seconds, \ie~$N=6000$. On each of these pieces of signal, we implement the forecasting method introduced in section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. However, the TBATS extension method in not implemented here because of its excessive computing time. The extensions of $7$ seconds-long on each boundary, corresponding to $L =700$. Thus, in order to catch slowly varying dynamical behaviors, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The average and the median of the MSE~\eqref{eq:mse} with respect to the simulations are given in Table~\ref{tab:mse}. The averaged MSE of {\sf SigExt} method is overwhelmingly higher than the averaged MSE of the other methods. This result is caused by the presence of outliers in some extensions by {\sf SigExt}. Indeed, the value median of the MSE is of the same order than those of the other methods. These outliers are probably due to the presence of pulse in the respiratory signal. {\color{red}[HT: plot histogram. Show a failure case.]} Those cases make the AHM model temporarily irrelevant, and break the validity of the linear dynamical model~\eqref{eq:dyn.model} used to extend the signal. We see that {\sf SigExt} is less robust to the vanishing of oscillations that the GPR or EDMD extensions.

\begin{table}
\centering
\caption{Respiratory signal. MSE of the respiratory signal extensions.}
\begin{tabular}{|c||c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE} \\
   \cline{2-3}
      & Mean & Median\\
   \hhline{|=#=|=|}
   {\sf SigExt} & $1.704 \times 10^{0}$ & $7.346 \times 10^{-1}$ \\
   \hline
   Symmetric & ${\color{red}X}$ & ${\color{red}X}$ \\
   \hline
   EDMD & $2.172\times 10^{-2}$ & $7.349\times 10^{-1}$ \\
   \hline
   GPR & $4.476\times 10^{-2}$ & $1.998\times 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:mse}
\end{table}

Then, we apply the boundary effects reduction method {\sf BoundEffRed} (Algorithm~\ref{alg:boundary}) on these extensions for diverse TF representations: STFT, SST, RS, as well as concentration of frequency and time (ConceFT), an generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:otd.tho}, the results are compared in terms of OTD. {\color{red}[HT: make clear where you do the comparison. On the whole signal or only near the boundary.]} They are also compared with the strategy consisting in a zero-padding extension of the signal. Even though the performance of the forecasting Algorithm~\ref{alg:extension} is somehow moderate, the boundary effects can be reduced dramatically on the TF representations. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length we use to evaluate the STFT is of 1500 samples. To prevent the STFT from being sensitive to the boundary effects, we set $L=750$. In this way, when evaluating the spectral content of the signal near its boundaries, the analysis is not limited by a lack of information all along the window support. From now on, all results are given for $L$ at equal to the half of the width of the window used in the TF transform.

\begin{table}
\centering
\caption{Respiratory signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|>{\centering}m{35pt}||c|c|c|c|}
  \hline
   \multirow{2}{35pt}{\centering Extension method} & \multicolumn{4}{c|}{TF Representation} \\
   \cline{2-5}
      & STFT & SST & RS & ConceFT \\
   \hhline{|=#=|=|=|=|}
   Without extension & $2.16\!\times\! 10^{-2}$ & $5.26\!\times\! 10^{-3}$ & $3.07\!\times\! 10^{-2}$ & $1.41\!\times\! 10^{-2}$ \\
   \hline
   {\sf SigExt} & $1.72\!\times\! 10^{-2}$ & $4.00\!\times\! 10^{-3}$ & $2.43\!\times\! 10^{-2}$ & $1.12\!\times\! 10^{-2}$ \\
   \hline
   Symmetric & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$\\
   \hline
   EDMD & $1.75\!\times\! 10^{-2}$ & $4.23\!\times\! 10^{-3}$ & $2.45\!\times\! 10^{-2}$ & $1.09\!\times\! 10^{-2}$ \\
   \hline
   GPR & $1.84\!\times\! 10^{-2}$ & $4.10\!\times\! 10^{-3}$ & $2.48\!\times\! 10^{-2}$ & $1.24\!\times\! 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:otd.tho}
\end{table}

@@

We divide the signal into 32-second long pieces, and apply Algorithm~\ref{alg:boundary} on each piece. We provide in Table~\ref{tab:otd.ppg} the OTD to the optimal TF representation averaged over the signals. For all the considered TF representations, the results clearly shows that our algorithm reduce the influence of the boundary effects. This highlights the ability of our approach to limit the distortion due the boundary effects and provide a more accurate representations. Even though the {\sf SigExt} extension yields TF representations slightly more sensitive to boundary effects than the extensions given by EDMD or GPR, it is the only technique that allows a real-time implementation.

\begin{table}
\centering
\caption{PPG signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Extension method} & \multicolumn{3}{c|}{TF Representation} \\
   \cline{2-4}
      & STFT & SST & RS\\
   \hhline{|=#=|=|=|}
   Without extension & $2.52\times 10^{-2}$ & $9.41\times 10^{-2}$ & $1.03\times 10^{-1}$ \\
   \hline
   {\sf SigExt} & $1.22\times 10^{-3}$ & $7.31\times 10^{-2}$ & $1.11\times 10^{-1}$ \\
   \hline
   Symmetric & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$ \\
   \hline
   EDMD & $9.83\times 10^{-4}$ & $5.59\times 10^{-2}$ & $9.80\times 10^{-2}$ \\
   \hline
   GPR & $1.14\times 10^{-3}$ & $7.02\times 10^{-2}$ & $1.07\times 10^{-1}$ \\
   \hline
\end{tabular}
\label{tab:otd.ppg}
\end{table}

On the bottom panel of Fig.~\ref{fig:ex.intro}, we display the SST resulting from the {\sf BoundEffRed} strategy, applied to the same portion of PPG than what is used to display Fig.~\ref{fig:ppg}. We clearly observe an improvement of the quality of the SST near boundaries. Indeed, the blurring visible when zooming on the right boundary of the SST has almost vanished. The real-time tracking of the instantaneous frequencies contained in the measured signal is therefore largely facilitated.

>>>>

On Fig.~\ref{fig:res.noise.sine}, we display the experimental variance $\gamma_{\xp}[N-1+\ell,N-1+\ell]$ for each value of $\sigma$. The associated theoretical asymptotic forecasting variance~\eqref{eq:cov.error.2} is also displayed in solid line. As expected, this result highlights the fact that the forecasting variance increases linearly with respect to $\sigma^2$. Surprisingly, this result shows that the forecasting variance {\em slightly} decreases with $\ell$, what is counterintuitive. 
%
{\color{red} [HT: need a more convincing explanation.] } 
%It should be noted that, contrary to what expression~\eqref{eq:cov.error.2} suggests, smaller values of $\sigma$ do not cause a decrease of the experimental variance, but an increase. This comes from the fact that when $\sigma$ is small, the matrix $\bX\bX^T$ becomes ill-conditioned. The calculation of the forecasting matrix $\tilde\bA$ in~\eqref{eq:lse} is then strongly disturbed.

\begin{figure}
%\includegraphics[width=.24\textwidth]{biasNoiseSine.eps}
\includegraphics[width=.48\textwidth]{VarianceNoiseSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $\sigma$.}
\label{fig:res.noise.sine}
\end{figure}

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary over time. The deterministic component $\bz$ of the random vector $\bx$ (constructed following the model~\eqref{eq:model.noise}) takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\[-1mm]
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the additive noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$. Numerically, we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

To quantify the global quality (\ie~not depending on $\ell$) of the forecasting approaches, we evaluate the Experimental Mean Square Error $\mathrm{MSE_{xp}}(\tilde\bx)$ of the forward forecast extended signals, namely:
\begin{align}
\label{eq:mse}
\mathrm{MSE_{\xp}}(\tilde\bx) &= \dfrac1{L}\|\tilde\bx -\bx^\mathrm{ext}\|^2 \\[-1mm]
\nonumber
&=\! \dfrac1{L}\sum_{\ell=1}^L \bmu_{\xp}[N\!-\!1\!+\!\ell]^2 \!+\! \bgamma_{\xp}[N\!-\!1\!+\!\ell,N\!-\!1\!+\!\ell] .
\end{align}
where $\bx^\mathrm{ext}$ is the ground-truth extended signal, that is: $\bx^\mathrm{ext} = \begin{pmatrix}\bx[-L] & \cdots & \bx[N-1+L] \end{pmatrix}$. Then, as long as the bias $\bmu[N-1+\ell]$ and the variance $\bgamma[N-1+\ell,N-1+\ell]$ of the forecasting estimator remain small for all $\ell$, the MSE takes small values either. Corresponding results are given in Table~\ref{tab:mse.sine}. They show that the naive extension we propose gives satisfying results, in particular in comparison with the point-symmetric extension. Besides, even though the other more sophisticated methods, like GPR, give MSE values that have a slightly smaller standard deviation, these methods are substantially limited by the computing time they require, which prevent them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimize the trade-off between the forecasting quality and the computing time. %That is why, it is implemented in our algorithm for the reduction of boundary effects.

@@

\subsubsection{Metrics}
The quality of the boundary effects reduction is evaluated directly on the TF representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$, defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation}
\ccF_N^\mathrm{opt}(\bx) = \cR\left( \ccF_{N+2L}(\bx^\mathrm{ext}) \right) \ .
\label{eq:opt.TFR}
\end{equation}

In the aim of comparing the different techniques, we use a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given TF representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following probability density function: $p_\ccQ^t(\xi) = \left. |\ccQ(\xi,t)|^2 \middle/ \int_\RR |\ccQ(\nu,t)|^2\dd\nu \right.$. At each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$. The \textit{optimal transport distance} quantifies the proximity between the estimated and actual instantaneous frequencies while favoring the sparsity of the estimated TF representation. That is why, the performance index $D(\ccQ)$ for the reduction of boundary effects of a given TF representation $\ccQ$ is given by the averaged ratio between the optimal transport distance to the optimal TF representation~\eqref{eq:opt.TFR} and the optimal transport distance of to the optimal TF representation, that is:
\begin{equation}
D(\ccQ) = \dfrac1{|I|}\bigintss_I\ \frac{d_{t}\left(\ccQ,\ccF_N^\mathrm{opt}\right)}{d_{t}\left(\ccF_N,\ccF_N^\mathrm{opt}\right)}\ \dd t\ .
\label{eq:index.perf}
\end{equation}
Thus, $D(\ccQ)<1$ means a reduction of the boundary effects. Let us evaluate the quality of the boundary effects reduction on biomedical signals.

\subsubsection{Respiratory signal}
We first consider a 6 hours 20 minutes-long respiratory signal. This signal is sampled at $\fs=100$~Hz. A zoom on a small portion of the signal is displayed in Fig.~\ref{fig:tho}.

@@

From that large signal, we build a dataset of $378$ non-overlapping signals of 60 seconds, \ie~$N=6000$. On each of these pieces of signal, we implement the forecasting method introduced in section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. However, the TBATS extension method in not implemented here because of its excessive computing time. The extensions of $7$ seconds-long on each boundary, corresponding to $L =700$. Thus, in order to catch slowly varying dynamical behaviors, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The resulting MSE~\eqref{eq:mse} averaged with respect to the simulations are given in Table~\ref{tab:THO}. The MSE of {\sf SigExt} is higher than the MSE of the other methods. This is mainly caused by the presence of a few segments of the whole signal involving complicated behaviors, that are unpredictable via a too simple dynamical model like~\eqref{eq:dyn.model}. The left of Fig.~\ref{fig:THO.failure} illustrates one of those cases, where {\sf SigExt} fails to catch the fast varying dynamic of the instantaneous amplitude to satisfactorily forecast the signal. The EDMD an Symmetric extensions are more robust to those situations, as shown on this example and in Table~\ref{tab:THO}. Nevertheless, {\sf SigExt} provides a sufficiently relevant extension to give TF representations sparingly affected by boundary effects. On the right of Fig.~\ref{fig:THO.failure}, we display comparison between the right boundary of SST of the same segment of signal (top-right), and its boundary-free SST obtained after the {\sf SigExt} forecasting (bottom-right). The extension of the instantaneous frequency visible on the right side of the image, illustrates the reduction of boundary effects produced despite an inaccurate signal forecasting.

We then apply {\sf BoundEffRed} (\ie~Algorithm~\ref{alg:boundary}) for diverse TF representations: STFT, SST, RS, as well as concentration of frequency and time (ConceFT), a generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:THO}, we give the averaged performance index~\eqref{eq:index.perf}, evaluated the whole TF representations (including  boundaries). Even though {\sf SigExt} performs somehow moderately, the boundary effects are dramatically reduced on the TF representations, in the same order of magnitude than with the forecastings given by EDMD or GPR. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length used to evaluate the STFT is of $1500$ samples. To prevent the STFT from being sensitive to the boundaries, we set $L=750$. In this way, the evaluation of the spectral content of the signal near its boundaries is not limited by a lack of information all along the window support. From now on, all results are given for $L$ equal to the half of the width of the window used in the TF transform.

\begin{table}
\centering
\caption{Performance of the extension methods on a respiratory signal.}
\begin{tabular}{|c||c||c|c|c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multirow{2}{35pt}{\centering Averaged MSE} & \multicolumn{4}{c|}{Averaged performance index $D$} \\
   \cline{3-6}
      & & STFT & SST & RS & ConceFT \\
   \hhline{|=#=#=|=|=|=|}
   {\sf SigExt} & $0.046$ & $0.705$ & $0.698$ & $0.694$ & $0.776$ \\
   \hline
   Symmetric & $0.039$ & $1.280$ & $0.984$ & $9.063$ & $1.295$ \\
   \hline
   EDMD & $0.022$ & $0.757$ & $0.635$ & $0.701$ & $0.793$ \\
   \hline
   GPR & $0.045$ & $0.817$ & $0.793$ & $0.727$ & $0.876$ \\ 
   \hline
\end{tabular}
\label{tab:THO}
\end{table}

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{THOfailure.eps}
\caption{Extensions of a segment of the respiratory signal (left) where {\sf SigExt} is outperformed by the EDMD and Symmetric extensions. Corresponding SST (top-right) and boundary-free SST obtained with {\sf SigExt} (bottom-right).}
\label{fig:THO.failure}
\end{figure}

@@

We divide the signal into 32-second long segments, and apply Algorithm~\ref{alg:boundary} on each piece. We provide in Table~\ref{tab:otd.ppg} the performance index $D$ of the boundary-free TF representations averaged over the signals. For all the considered TF representations, the results clearly shows that our algorithm reduces the boundary effects. %This highlights the ability of our approach to limit the distortion due the boundary effects and provide a more accurate representations. 
Even though, on this signal, the {\sf SigExt} extension yields TF representations slightly more sensitive to boundary effects than the extensions given by EDMD or GPR, it is the only technique that allows a real-time implementation.

\begin{table}
\centering
\caption{PPG signal: averaged performance index of the boundary-free TF representations for different extensions.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Extension method} & \multicolumn{3}{c|}{Performance index $D$} \\
   \cline{2-4}
      & STFT & SST & ConceFT\\
   \hhline{|=#=|=|=|}
%   Without extension & $2.52\times 10^{-2}$ & $9.41\times 10^{-2}$ & $1.03\times 10^{-1}$ \\
%   \hline
   {\sf SigExt} & $0.774$ & $0.048$ & $X$ \\
   \hline
   Symmetric & $0.818$ & $0.059$ & $X$ \\
   \hline
   EDMD & $0.566$ & $0.038$ & $X$ \\
   \hline
   GPR & $0.735$ & $0.045$ & $X$ \\
   \hline
\end{tabular}
\label{tab:otd.ppg}
\end{table}

On the bottom-right panel of Fig.~\ref{fig:ex.intro}, we display the SST resulting from the {\sf BoundEffRed} strategy, applied to the portion of PPG displayed on Fig.~\ref{fig:ppg}. We clearly observe an improvement of the quality of the SST near boundaries. Indeed, the blurring visible when zooming on the right boundary of the SST has almost vanished. The real-time tracking of the instantaneous frequencies contained in the measured signal is therefore largely facilitated.
